---
title: "Problem set 4"
subtitle: "PPHA 31102 Statistics for Data Analysis II: Regressions"
author: "Jay Ballesteros"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
geometry: margin=1in
fontsize: 12pt
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \fvset{fontsize=\small}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(AER)
```

**Study Group Members:** [List names here]

---

# 1. Trade Policy (8 points)

```{r load-trade-data}
# Load the trade policy data
trade <- read.csv("trade_policy.csv")
```

## Question 1.1 (4 points)

*Use a difference-in-differences design to estimate the impact of the tariff on exports of air conditioners. Regress air_conditioners_pc on post, treat, and the interaction of post and treat. Which coefficient reveals the change in exports after the tariff was introduced? Explain and interpret the result.*

```{r q1-1}
# DiD regression

```

**Answer:** [Your interpretation here]

## Question 1.2 (2 points)

*A friend is suggesting that you should consider taking the log of your outcome, air_conditioners_pc. What are some potential concerns regarding this suggestion? Describe two.*

**Answer:** [Your answer here]

## Question 1.3 (2 points)

*The difference-in-differences design is valid when treatment and control have common trends. What is a common trend? Write down the definition. Next, give an example (about this case) when this assumption is likely violated.*

**Answer:** [Your answer here]

---

# 2. The Oregon Health Insurance Experiment, Revisited (16 points)

```{r load-ohie-data}
# Load the OHIE data
ohie <- read.csv("OHIE.csv")
```

## Question 2.1 (2 points)

*Consider the variable ever_medicaid. Did everyone who won the lottery take up Medicaid? Calculate the mean of ever_medicaid among those who won the lottery. Are there some people who lost the lottery who take up Medicaid? Calculate the mean of ever_medicaid for those that lost the lottery. Is the causal effect of winning the lottery likely to be the same as the causal effect of actually receiving Medicaid?*

```{r q2-1}
# Mean of ever_medicaid among lottery winners

# Mean of ever_medicaid among lottery losers

```

**Answer:** [Your interpretation here]

## Question 2.2 (2 points)

*Estimate the "First-stage" regression: ever_medicaid = α0 + α1*treated + α2*numhh_list + e. Interpret the regression coefficient on α1.*

```{r q2-2}
# First-stage regression

```

**Answer:** [Your interpretation here]

## Question 2.3 (3 points)

*Is treated likely to be a valid instrument for ever_medicaid in terms of the three key assumptions: instrument relevance, independence assumption, and the exclusion restriction?*

**Answer:** [Your answer here]

## Question 2.4 (2 points)

*Calculate and report the ratio of the reduced form to the first stage for the outcome count_visit_dr. Note: use the same observations (exclude rows with missing count_visit_dr).*

```{r q2-4}
# Subset to non-missing count_visit_dr

# Reduced form regression

# First-stage regression (same sample)

# Calculate ratio

```

**Answer:** [Your interpretation here]

## Question 2.5 (3 points)

*Manually estimate the two stages of 2SLS. How does your result compare to the previous estimate? Is the IV estimate statistically significant?*

```{r q2-5}
# (a) First-stage and predicted values

# (b) Second-stage regression

```

**Answer:** [Your interpretation here]

## Question 2.6 (2 points)

*Re-estimate the 2SLS model including the full set of controls (numhh_list, female, age, age^2, race_white, hs_degree, college_degree, health_baseline). Compare the coefficient on ever_medicaid and the standard error to the results in part (5).*

```{r q2-6}
# Create age squared variable

# 2SLS with full controls

```

**Answer:** [Your comparison here]

## Question 2.7 (2 points)

*For each endline outcome, run the IV regressions including controls using ivreg(). Fill in the table and discuss findings.*

```{r q2-7}
# IV regression for visit_dr

# IV regression for visit_er

# IV regression for out_of_pocket_spend

# IV regression for health_score

# IV regression for happy

```

| Endline outcome    | (1) IV estimate | (2) S.E. |
|--------------------|-----------------|----------|
| visit_dr           |                 |          |
| visit_er           |                 |          |
| out_of_pocket_spend|                 |          |
| health_score       |                 |          |
| happy              |                 |          |

**Answer:** [Your discussion here]

